Relational Parametricity for Algebraically Indexed Types

### 2012-06-05 paper/geometry-twocolumn.tex
Redo the geometry section to just cover the origin invariance

@@

FIXME: vectors are not invariant under change of origin, so they are
represented by the relation $R_0$, which is just equality. Bxut they,
and points are invariant under change of basis. However, as we shall
see below, not all operations are invariant under all changes of
basis. In particular the dot product is only invariant under
orthogonal transformations.

@@

In the last section, we saw how Reynolds' notion of preservation of
relations can be used to define when it is the case that programs
manipulating representatives of points and vectors make geometric
sense. Reynolds further insight was to see how preservation of
relations can automatically be established by following a typing
discipline. In Reynolds' case, the typing discipline was the
polymorphic $\lambda$-calculus where new types can be constructed by
universal quantification over all types. Universal type quantification
is commonly written $\forall \alpha. A$, which denotes a type that may
be instantiated with any other type $B$ by replacing the type variable
$\alpha$ with $B$ in $A$.

In terms of relations, Reynolds interprets universal quantification
over types as quantification over binary relations between denotations
of types. In the statements of geometric invariance that we stated in
the previous section (e.g., \statementref{eq:f-invariant-frame}), we
did not quantify over all relations. Instead, we quantified over all choices of origin

elements of a geometric group such as $\GA(2)$ or $\GL(2)$, and used
those elements to select relations from the families $R_{\texttt{pt}}$
and $R_{\texttt{vec}}$. Driven by this observation, we introduce
quantification over elements of geometric groups into the language of
types. We use the notation $\forall X \mathord: G. A$ for
quantification over elements of geometric groups, where $G$ is the
name of some geometric group, and $X$ is a variable name standing for
an element of $G$. We refer to $G$ as the \emph{sort} of $X$. As we
proceed, we will introduce the sorts representing geometric groups
that we will be using in addition to $\GA(2)$ and $\GL(2)$.

The sorts $\GA(2)$ and $\GL(2)$ represent groups, so we can combine
their elements using the usual group operations. We write operations
multiplicatively, using juxtaposition ($e_1e_2$) for multiplication,
$e^{-1}$ for inverse and $1$ for the unit. We overload the same
notation to stand for the group operations in both $\GA(2)$ and
$\GL(2)$. We also regard expressions built from variables and the
group operations up to the group axioms. So, for example,
$F_1(F_2F_3)$ and $(F_1F_2)F_3$ are to be regarded as equivalent.

Our language of types also includes the common type constructors for
function types, $A \to B$, sum types $A + B$, unit type
$\mathsf{Unit}$ and tuple types $A \times B$. We also assume a
primitive type $\mathsf{real}$, used to represent scalars. As we noted
at the start of the previous section, the central data structure in
geometric applications is the tuple of real numbers for representing
points and vectors. However, we cannot simply express this as the type
$\mathsf{real} \times \mathsf{real}$ because it will not have the
correct relational interpretation. (Two elements of type
$\mathsf{real}$ will be related if and only if they are equal, and
hence two elements of $\mathsf{real} \times \mathsf{real}$ will be
related if and only if they are equal, by Reynolds' definition of the
relational interpretation of tuple types.) Therefore, we introduce a
new type parameterised by expressions $e$ of sort $\GA(2)$ to
represent points (and later, vectors): $\mathsf{vec}\langle e
\rangle$. We have used the name $\mathsf{vec}$, even though we have
taken pains to separate geometric points and vectors, to recall the
computer science notion of vector as a sequence of values of
homogeneous type which has a known length (in this case two).

@@

Equipped with quantification over the group $\GA(2)$ of invertible
affine maps, and the parameterised types $\mathsf{vec}\langle \cdot
\rangle$, we can now express geometric invariance properties as
types. In the previous section, we gave examples of functions $f :
\mathbb{R}^2 \times \mathbb{R}^2 \to \mathbb{R}^2$ that have the
property that they preserve all invertible affine maps. We can now
express this property as a type:
\begin{displaymath}
  \mathrm{f} : \forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \times \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle
\end{displaymath}
We use the variable name $F$ for elements of $\GA(2)$ to suggest the
intuition that this typing indicates that $\mathrm{f}$ takes a pair of
points in the $F$rame $F$ and returns a point in the same frame.

\subsection{Affine Combination of Points}
\label{sec:affine-combination}

As we stated in the previous section, the primitive operation of
affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
\lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
affine combination can be interpreted as interpolation between the
points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
combination of points to our calculus, with the following typing and
intended denotation.
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affineCombination} :\\
    \hspace{1.5cm}\forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \\
    \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ t\ (x_2,y_2) = \\
    \hspace{3cm}((1-t)x_1 + tx_2, (1-t)y_1 + ty_2)
  \end{array}
\end{displaymath}
By defining the primitive function $\mathrm{affineCombination}$ to
take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
we are taking the affine combination of two representatives of points,
and not just an arbitrary linear combination. Note that the intended
denotation in the second line does not take any argument corresponding
to the universally quantified $F$ in the type. In our calculus,
geometric group expressions are erased at runtime.

% In Linear Algebra notation, we could have written the right-hand side
% as $(1-t)\vec{v_1} + t\vec{v_2}$, where $\vec{v_1} = (x_1,y_1)$ and
% $\vec{v_2} = (x_2,y_2)$.

\begin{example}
  With just the single primitive operation of affine combination, we
  can already express some useful programs. For example, the
  evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves with
  two endpoints and a single control point) can be expressed using the
  affine combination primitive and three steps of De Casteljau's
  algorithm:
  \begin{displaymath}
    \begin{array}{l}
      \mathrm{quadraticB\acute{e}zier} :\\
      \hspace{0.5cm}\forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \\
      \mathrm{quadraticB\acute{e}zier}\ [F]\ p_0\ p_1\ p_2\ t = \\
      \quad
      \begin{array}{l@{\hspace{-0.002cm}}l}
        \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ t\ p_1) \\
        & t \\
        & (\mathrm{affineCombination}\ [F]\ p_1\ t\ p_2)
      \end{array}
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
  application $\mathrm{quadraticB\acute{e}zier}\ p_0\ p_1\ p_2\ t$ for
  $t \in [0,1]$ gives us the point on the curve at ``time'' $t$. Note
  that the type of $\mathrm{quadraticB\acute{e}zier}$ tells us
  immediately that it preserves all invertible affine maps.
  %FIXME: forward ref to ``free theorem''.
\end{example}

@@

Affine combination only provides us with an operation on points. For
the purposes of computational geometry, we are also interested in
operations on the vectors representing offsets between points. We now
examine the correct types to assign to the vector space operations of
addition of vectors, negation of vectors, multiplication by a scalar
and the zero vector.

In our discussion in the previous section, we observed that vector
operations are those that are preserved by change-of-basis operations
-- i.e.~under the action of elements of the group $\GL(2)$ of
invertible linear maps. Since $\GL(2)$ is a subgroup of $\GA(2)$ we
may form the type $\mathsf{vec}\langle e \rangle$ when $e$ is of sort
$\GL(2)$ as well as $\GA(2)$. The obvious type assignment for vector
addition is now the following:
\begin{displaymath}
  (+) : \forall B\mathord:\GL(2).\ \mathsf{vec}\langle B \rangle \to \mathsf{vec}\langle B \rangle \to \mathsf{vec}\langle B \rangle
\end{displaymath}
(we write binary operators intended to be used infix in parentheses
when not appearing in infix position, following the Haskell
syntax.) The use of $B$ for names of elements of $\GL(2)$ is meant to
suggest that the type $\mathsf{vec}\langle B \rangle$ denotes vectors
with respect to the $B$asis $B$.

However, it is possible to give vector addition a more precise type
that describes its effect on the chosen origin as well as its
invariance under change of basis. We now introduce a new sort
$\Transl(2)$ of translations. Translations form an abelian group,
which we write additively, using $+$,$-$ and $0$, and we use lower
case $t$, $t_1$, etc.~for $t$ranslation variables. Invertible affine
maps are constructed from invertible linear maps and translations, so
given a $B : \GL(2)$ and a $t : \Transl(2)$, we can form an invertible
affine map $B \ltimes t : \GA(2)$. Conversely, given $F : \GA(2)$, we
have an invertible linear map $\pi_L(F) : \GL(2)$, and a translation
$\pi_T(F) : \Transl(2)$. Thinking of elements of $\GA(2)$ as frames,
$\pi_L$ returns the basis of the frame, and $\pi_T$ returns the
origin.

We can now give vector addition a more general type, which
demonstrates how adding vectors has an effect on translations:
\begin{displaymath}
  \begin{array}{l}
    (+) : \forall B \mathord: \GL(2), t_1, t_2 \mathord: T(2).\ \\
    \quad\quad\quad\quad\mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2 \rangle \to \mathsf{vec}\langle B \ltimes (t_1 + t_2)\rangle
\end{array}
\end{displaymath}
We can also negate vectors, yielding a vector which points in the
opposite direction. Negation is invariant under invertible linear maps
but negates translations:
\begin{displaymath}
  \mathrm{negate} : \forall B \mathord: \GL(2), t \mathord: T(2).\ \mathsf{vec}\langle B \ltimes t\rangle \to \mathsf{vec}\langle B\ltimes(-t)\rangle
\end{displaymath}
With addition and negation of vectors, we can define the derived
operation of vector subtraction with the following type:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall B \mathord:\GL(2), t_1,t_2 \mathord:T(2).\ \\
    \hspace{1cm}\mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2\rangle \to \mathsf{vec}\langle B \ltimes (t_1 - t_2)\rangle \\
    (-)\ [B]\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
\end{array}
\end{displaymath}

Given two points in the same frame---i.e.~two values of type
$\mathsf{vec}\langle F \rangle$---we can compute their offset, which
is a vector expressed with respect to the basis part of $F$, as a
special case of vector subtraction:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle \pi_L(F) \ltimes 0 \rangle \\
    \mathrm{offset}\ [F]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
The subtraction operation in the body of $\mathrm{offset}$ is
instantiated with $B = \pi_L(F)$, $t_1 = \pi_T(F)$ and $t_2 =
\pi_T(F)$. The translation $\pi_T(F)$ cancels itself out, and we are
left with a vector in the basis $\pi_L(F)$ of the frame $F$.

Similarly, the type of the vector addition operation can be
specialised to the case of moving a point by an offset represented by
a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle \pi_L(F) \ltimes 0 \rangle \to \mathsf{vec}\langle F \rangle \\
    \mathrm{moveBy}\ [F]\ p\ v = p + v
  \end{array}
\end{displaymath}
The addition operation in the body of $\mathrm{moveBy}$ is
instantiated with $B = \pi_L(F)$, $t_1 = \pi_T(F)$ and $t_2 =
0$. Since $t_1 + t_2 = \pi_T(F) + 0 = \pi_T(F)$ in this case, we are
left with a value of type $\mathsf{vec}\langle \pi_L(F) \ltimes
\pi_T(F) \rangle$, which is equivalent to the type
$\mathsf{vec}\langle F \rangle$.

The types we assign to the remaining vector space primitives, the zero
vector and multiplication by a scalar, do not describe any interesting
effect on translations%\footnote{but they could...}
. The zero vector has the following polymorphic type, indicating that
for any basis $B$, there is a zero vector.
\begin{displaymath}
  \mathrm{0} : \forall B \mathord: \GL(2).\ \mathsf{vec}\langle B \ltimes 0 \rangle
\end{displaymath}
Intuitively, there cannot be a zero vector for any frame $F$, because
this would entail picking a privileged origin. For multiplication of a
vector by a scalar, we assign the following type:
\begin{displaymath}
  (*) : \forall B \mathord: \GL(2).\ \mathsf{real} \to \mathsf{vec}\langle B \ltimes 0 \rangle \to \mathsf{vec}\langle B \ltimes 0 \rangle
\end{displaymath}
This type indicates that scaling a vector is preserved by change of
basis.

\begin{example}
  The vector space primitives we have defined in this section, along
  with the geometric invariance properties that follow from their
  types, allow us to establish a useful type isomorphism. Consider
  functions with types conforming to the following schema:
  \begin{displaymath}
    \tau_{n,A} \isDefinedAs \forall F\mathord:\GA(2).\ \mathsf{vec}\langle F \rangle^{n+1} \to A[F]
  \end{displaymath}
  We take $n$ to be a natural number, and $A^n$ for some type $A$
  denotes the $n$-ary tupling of $A$ with itself: i.e.~$A \times
  ... \times A$, with $A$ repeated $n$ times. The notation $A[-]$
  denotes a type with a ``hole'' that can be filled with an
  expression. In this case, we require that $A[e]$ be a well-formed
  type when $e$ has sort $\GA(2)$.

  Just by looking at the types $\tau_{n,A}$, we know that their
  inhabitants will be invariant under change of frame, due to the
  quantification over all $F$ in $\GA(2)$. Therefore, we may as well
  pick one of the input points as the origin and assume that all the
  other points are defined with respect to this chosen origin. This is
  similar to the common mathematical step of stating that, ``without
  loss of generality'', we may pick some point in a description of a
  problem to be the origin, as long as the problem statement is
  invariant under translation.

  The types $\tau_{n,A}$ are isomorphic to the corresponding types
  $\sigma_{n,A}$, defined below. Inhabitants of the types
  $\sigma_{n,A}$ take one fewer points (now seen as vector offsets
  from some origin), but are invariant only under change of basis.
  \begin{displaymath}
    \sigma_{n,A} \isDefinedAs \forall B\mathord:\GL(2).\ \mathsf{vec}\langle B \rangle^n \to A[B \ltimes 0]
  \end{displaymath}
  The isomorphism between these two families of types can be witnessed
  by the following two functions.
  \begin{displaymath}
    \begin{array}{l}
      i_{n,A} : \tau_{n,A} \to \sigma_{n,A} \\
      i_{n,A}\ f\ [B]\ (p_1, p_2, ..., p_n) = f\ [B \ltimes 0]\ (0, p_1, ..., p_n) \\
      \\
      i^{-1}_{n,A} : \sigma_{n,A} \to \tau_{n,A} \\
      i^{-1}_{n,A}\ g\ [F]\ (p_0, ..., p_n) = g\ [\pi_L(F)]\ (p_1-p_0, ..., p_n-p_0)
    \end{array}
  \end{displaymath}
  The direction defined by the function $i_{n,A}$ treats the inputs
  $p_1,...,p_n$ as vector offsets from the origin $0$. The direction
  witnessed by $i^{-1}_{n,A}$ picks the first point to act as the
  origin, and uses the operation of vector subtraction we defined
  above to turn each of the other points into offsets from this
  point. Note that this isomorphism is not unique: for each $n$ we can
  pick any of the $n+1$ inputs to act as the distinguished
  origin. Thus the families of types $\tau_{n,A}$ and $\sigma_{n,A}$
  are isomorphic, but not uniquely isomorphic.
\end{example}

>>>>

\textbf{FIXME:} talk about the standard way of separating the two
concepts using abstract types.

@@

% FIXME: vectors are not invariant under change of origin, so they are
% represented by the relation $R_0$, which is just equality. But they,
% and points are invariant under change of basis. However, as we shall
% see below, not all operations are invariant under all changes of
% basis. In particular the dot product is only invariant under
% orthogonal transformations.

@@

We have seen how Reynolds' notion of preservation of relations can be
used to define when it is the case that programs manipulating
representatives of points and vectors make geometric sense. Reynolds
further insight was to see how preservation of relations can
automatically be established by following a typing discipline. In
Reynolds' case, the typing discipline was the polymorphic
$\lambda$-calculus where new types can be constructed by universal
quantification over all types. Universal type quantification is
commonly written $\forall \alpha. A$, which denotes a type that may be
instantiated with any other type $B$ by replacing the type variable
$\alpha$ with $B$ in $A$.

In terms of relations, Reynolds interprets universal quantification
over types as quantification over binary relations between denotations
of types. In the statements of geometric invariance that we stated in
the previous section (e.g., \statementref{eq:f-invariant-frame}), we
did not quantify over all relations. Instead, we quantified over all
choices of origin and used the choice to select a relation from the
family $\{R_{\vec{o}}\}$. Driven by this observation, we introduce
quantification over choice of origin into the language of types.  We
use the notation $\forall t \mathord: T_2. A$ for quantification over
all 2-dimensional translations (i.e.~choices of origin) $t$. We refer
to $T_2$ as the \emph{sort} of $t$.

% As we proceed, we will introduce the sorts representing geometric
% groups that we will be using in addition to $\GA(2)$ and $\GL(2)$.

The sort $T_2$ represents an abelian group, so we can combine its
elements using the usual group operations. We write operations
additively, using $e_1 + e_2$ for the group operation, $-e$ for
inverse and $0$ for the unit.  We also regard expressions built from
variables and the group operations up to the group axioms. So, for
example, $e_1 + (e_2 + e_3)$ and $(e_1 + e_2) + e_3$ are to be
regarded as equivalent.

Our language of types also includes the common type constructors for
function types, $A \to B$, sum types $A + B$, unit type
$\mathsf{Unit}$ and tuple types $A \times B$. We also assume a
primitive type $\mathsf{real}$, used to represent scalars. As we noted
at the start of the previous section, the central data structure in
geometric applications is the tuple of real numbers for representing
points and vectors. However, we cannot simply express this as the type
$\mathsf{real} \times \mathsf{real}$ because it will not have the
correct relational interpretation. (Two elements of type
$\mathsf{real}$ will be related if and only if they are equal, and
hence two elements of $\mathsf{real} \times \mathsf{real}$ will be
related if and only if they are equal, by Reynolds' definition of the
relational interpretation of tuple types.) Therefore, we introduce a
new type parameterised by expressions $e$ of sort $T_2$ to represent
points (and later, vectors): $\mathsf{vec}\langle e \rangle$. We have
used the name $\mathsf{vec}$, even though we have taken pains to
separate geometric points and vectors, to recall the computer science
notion of vector as a sequence of values of homogeneous type which has
a known length (in this case two).

@@

Equipped with quantification over the group $T_2$ of invertible affine
maps, and the parameterised types $\mathsf{vec}\langle \cdot \rangle$,
we can now express origin invariance properties as types. In the
previous section, we discussed functions $f : \mathbb{R}^2 \times
\mathbb{R}^2 \to \mathbb{R}^2$ that have the property that they
preserve all changes of origin. We can now express this property as a
type:
\begin{displaymath}
  \mathrm{f} : \forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \times \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle
\end{displaymath}
FIXME: forward references...

% \subsection{Affine Combination of Points}
% \label{sec:affine-combination}

% As we stated in the previous section, the primitive operation of
% affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
% \lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
% affine combination can be interpreted as interpolation between the
% points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
% combination of points to our calculus, with the following typing and
% intended denotation.
% \begin{displaymath}
%   \begin{array}{l}
%     \mathrm{affineCombination} :\\
%     \hspace{1.5cm}\forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \\
%     \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ t\ (x_2,y_2) = \\
%     \hspace{3cm}((1-t)x_1 + tx_2, (1-t)y_1 + ty_2)
%   \end{array}
% \end{displaymath}
% By defining the primitive function $\mathrm{affineCombination}$ to
% take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
% we are taking the affine combination of two representatives of points,
% and not just an arbitrary linear combination. Note that the intended
% denotation in the second line does not take any argument corresponding
% to the universally quantified $F$ in the type. In our calculus,
% geometric group expressions are erased at runtime.

% % In Linear Algebra notation, we could have written the right-hand side
% % as $(1-t)\vec{v_1} + t\vec{v_2}$, where $\vec{v_1} = (x_1,y_1)$ and
% % $\vec{v_2} = (x_2,y_2)$.

% \begin{example}
%   With just the single primitive operation of affine combination, we
%   can already express some useful programs. For example, the
%   evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves with
%   two endpoints and a single control point) can be expressed using the
%   affine combination primitive and three steps of De Casteljau's
%   algorithm:
%   \begin{displaymath}
%     \begin{array}{l}
%       \mathrm{quadraticB\acute{e}zier} :\\
%       \hspace{0.5cm}\forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \\
%       \mathrm{quadraticB\acute{e}zier}\ [F]\ p_0\ p_1\ p_2\ t = \\
%       \quad
%       \begin{array}{l@{\hspace{-0.002cm}}l}
%         \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ t\ p_1) \\
%         & t \\
%         & (\mathrm{affineCombination}\ [F]\ p_1\ t\ p_2)
%       \end{array}
%     \end{array}
%   \end{displaymath}
%   For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
%   application $\mathrm{quadraticB\acute{e}zier}\ p_0\ p_1\ p_2\ t$ for
%   $t \in [0,1]$ gives us the point on the curve at ``time'' $t$. Note
%   that the type of $\mathrm{quadraticB\acute{e}zier}$ tells us
%   immediately that it preserves all invertible affine maps.
%   %FIXME: forward ref to ``free theorem''.
% \end{example}

@@

\textbf{FIXME}Affine combination only provides us with an operation on
points. For the purposes of computational geometry, we are also
interested in operations on the vectors representing offsets between
points. We now examine the correct types to assign to the vector space
operations of addition of vectors, negation of vectors, multiplication
by a scalar and the zero vector.

In our discussion in the previous section \textbf{?FIXME?}, we
observed that vector operations are not invariant under change of
origin. The obvious type assignment for vector addition is therefore:
\begin{displaymath}
  (+) : \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle
\end{displaymath}
(we write binary operators intended to be used infix in parentheses
when not appearing in infix position, following the Haskell
syntax.)

However, it is possible to give vector addition a more precise type
that describes its effect on the chosen origin. 
% We now introduce a new
% sort $\Transl(2)$ of translations. Translations form an abelian group,
% which we write additively, using $+$,$-$ and $0$, and we use lower
% case $t$, $t_1$, etc.~for $t$ranslation variables. Invertible affine
% maps are constructed from invertible linear maps and translations, so
% given a $B : \GL(2)$ and a $t : \Transl(2)$, we can form an invertible
% affine map $B \ltimes t : \GA(2)$. Conversely, given $F : \GA(2)$, we
% have an invertible linear map $\pi_L(F) : \GL(2)$, and a translation
% $\pi_T(F) : \Transl(2)$. Thinking of elements of $\GA(2)$ as frames,
% $\pi_L$ returns the basis of the frame, and $\pi_T$ returns the
% origin.
%
% We can now give vector addition a more general type, which
% demonstrates how adding vectors has an effect on translations:
\begin{displaymath}
  (+) : \forall t_1, t_2 \mathord: T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2 \rangle \to \mathsf{vec}\langle t_1 + t_2 \rangle
\end{displaymath}
We can also negate vectors, yielding a vector which points in the
opposite direction. Negation negates translations:
\begin{displaymath}
  \mathrm{negate} : \forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle -t \rangle
\end{displaymath}
With the primitive operations of addition and negation of vectors, we can define the derived
operation of vector subtraction with the following type:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall t_1,t_2 \mathord:T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2\rangle \to \mathsf{vec}\langle t_1 - t_2 \rangle \\
    (-)\ [B]\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
  \end{array}
\end{displaymath}

Given two points that have been constrained to be invariant with
respect to the same change of origin---i.e.~two values of type
$\mathsf{vec}\langle t \rangle$---we can compute their offset, which
is a vector expressed with respect to the null change of origin. The
offset operation is a special case of vector subtraction:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle 0 \rangle \\
    \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
Similarly, the type of the vector addition operation can be
specialised to the case of moving a point by an offset represented by
a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle t \rangle \\
    \mathrm{moveBy}\ [t]\ p\ v = p + v
  \end{array}
\end{displaymath}
The types we assign to the remaining vector space primitives, the zero
vector and multiplication by a scalar, do not describe any interesting
effect on translations%\footnote{but they could...}
. The zero vector has the following type.
\begin{displaymath}
  \mathrm{0} : \mathsf{vec}\langle 0 \rangle
\end{displaymath}
Intuitively, there cannot be a zero vector invariant under any change
of origin, because this would entail picking a privileged origin. For
multiplication of a vector by a scalar, we assign the following type,
which is again not invariant under change of origin.
\begin{displaymath}
  (*) : \mathsf{real} \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle
\end{displaymath}

\begin{example}
  The vector space operators and the properties that follow from their
  types allow us to establish a useful type isomorphism. Consider
  functions with types conforming to the following schema:
  \begin{displaymath}
    \tau_{n} \isDefinedAs \forall t\mathord:T_2.\ \mathsf{vec}\langle t \rangle^{n+1} \to \mathsf{real}
  \end{displaymath}
  We take $n$ to be a natural number, and $A^n$ for some type $A$
  denotes the $n$-ary tupling of $A$ with itself: i.e.~$A \times
  ... \times A$, with $A$ repeated $n$ times.
% The notation $A[-]$
%   denotes a type with a ``hole'' that can be filled with an
%   expression. In this case, we require that $A[e]$ be a well-formed
%   type when $e$ has sort $\GA(2)$.

  Just by looking at the types $\tau_{n}$, we know that their
  inhabitants will be invariant under change of origin, due to the
  quantification over all $t$ in $T_2$. Therefore, we may as well pick
  one of the input points as the origin and assume that all the other
  points are defined with respect to this chosen origin. This is
  similar to the common mathematical step of stating that, ``without
  loss of generality'', we may pick some point in a description of a
  problem to be the origin, as long as the problem statement is
  invariant under translation.

  The types $\tau_{n}$ are isomorphic to the corresponding types
  $\sigma_{n}$, defined below. Inhabitants of the types
  $\sigma_{n}$ take one fewer points (now seen as vector offsets
  from some origin), but are invariant only under change of basis.
  \begin{displaymath}
    \sigma_{n} \isDefinedAs \mathsf{vec}\langle 0 \rangle^n \to \mathsf{real}
  \end{displaymath}
  We demonstrate this isomorphism formally below in
  \autoref{sec:FIXME}.

  % The isomorphism between these two families of types can be witnessed
  % by the following two functions.
  % \begin{displaymath}
  %   \begin{array}{l}
  %     i_{n,A} : \tau_{n,A} \to \sigma_{n,A} \\
  %     i_{n,A}\ f\ [B]\ (p_1, p_2, ..., p_n) = f\ [B \ltimes 0]\ (0, p_1, ..., p_n) \\
  %     \\
  %     i^{-1}_{n,A} : \sigma_{n,A} \to \tau_{n,A} \\
  %     i^{-1}_{n,A}\ g\ [F]\ (p_0, ..., p_n) = g\ [\pi_L(F)]\ (p_1-p_0, ..., p_n-p_0)
  %   \end{array}
  % \end{displaymath}
  % The direction defined by the function $i_{n,A}$ treats the inputs
  % $p_1,...,p_n$ as vector offsets from the origin $0$. The direction
  % witnessed by $i^{-1}_{n,A}$ picks the first point to act as the
  % origin, and uses the operation of vector subtraction we defined
  % above to turn each of the other points into offsets from this
  % point. Note that this isomorphism is not unique: for each $n$ we can
  % pick any of the $n+1$ inputs to act as the distinguished
  % origin. Thus the families of types $\tau_{n,A}$ and $\sigma_{n,A}$
  % are isomorphic, but not uniquely isomorphic.
\end{example}

\begin{example}
  An inexpressible type
  \begin{displaymath}
    \forall t \mathord: T_2.\ \vec{t + t} \to \vec{t}
  \end{displaymath}
  FIXME: explain that it is possible to show that this type is
  uninhabited with the given operations. We show this in general 
\end{example}

\subsection{Generalising from Origin Invariance}

Points for consideration in terms of language design:
\begin{itemize}
\item Knowledge of the algebraic laws by the type checker are
  essential for type-checking the examples above
\item The relational interpretation gives us 
\end{itemize}
